Fused regression for multi-source network inference

### 2015-12-05 paper1.tex
hi

\title{Fused regression for network inference in multiple species}
\author{Kari Y. Lam, Zachary M. Westrick, Christian L. M\"{u}ller, Lionel Christiaen, Richard Bonneau}
\maketitle

\begin{abstract}
Learning gene regulatory networks is essential to our understanding of cell behavior and response to external stimuli. Methods for network inference from large datasets have been developed and used for a wide variety of species; however, most approaches consider the problem as independent in each species even though there is known to be conservation of gene regulation even in distantly related species. We introduce a method that allows data to be pooled across species in a principled way based on gene orthology and demonstrate that it improves network inference even when the orthology is incomplete. We further refine this method with an algorithm that attempts to learn the true conserved subnetwork from a larger set of potentially conserved interactions. We demonstrate the utility of our method not only in cross species network inference, but also in merging data sets and incorporating binding similarity predictions.
\end{abstract}

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable \cite{bonneau_predictive_2007, ciofani_validated_2012, carro_transcriptional_2010}. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist adequate data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from previous studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference. The core of both methods is the observation that biological information about the relatedness of two networks can be used to approximate which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our first method -- fused ridge -- method uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously learn the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different).

In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. We make the assumption that if a pair of TFs are orthologous to one another, then their regulatory effects on a pair of orthologous genes should be similar. This orthology information forms the basis of a large number of constraints on the network inference problems, which allow data from one organism to improve network inference on another organism. This very general framework for multi-species network inference can be extended to large numbers of distinct organisms, each contributing data with only partially overlapping sets of genes. We use our knowledge about the relationship between nodes across networks in order to constrain the network inference problem; this approach has many applications, from improving network inference of multiple species to combining data sets to leveraging knowledge about similarity in binding sites.

Existing multi species approaches often use orthology as a proxy for functional conservation \cite{penfold_inferring_2015, joshi_multi-species_2015, kashima_simultaneous_2009, zhang2010nearly}, or attempt to learn functional similarity via expression data \cite{gholami_cross-species_2010}. Orthology -- the measure of gene similarity we use -- can be approximated using readily identifiable sequence similarity. Sequence similarity is often a useful predictor of functional similarity \cite{wilson_assessing_2000}. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with evolutionary constraints created using orthology. However, we recognize that some genes may have taken on different functions and therefore may have new regulatory interactions. Gene duplications and neofunctionalization in multi-species network inference \cite{eisen_phylogenomics:_1998}, and changes in chromatin configuration in multi-cell line network inference \cite{li_role_2007}, are cases where enforcing fusion of constraints does not produce networks which reflect true differences in regulation between networks. These cases can be difficult to learn by simply comparing independently inferred networks; because network inference is typically underconstrainted \cite{marbach_revealing_2010-1}, there could be other plausible networks in which the coefficients being compared are similar. Our adaptive fusion method allows us to learn differences in networks predicted to be similar based on orthology or gene matching: because the constraints we impose favor weight configurations in which fused interaction weights are similar to one another, a solution in which a priori similar regulatory interactions are found to be different is strong evidence that they really are different. When differences in regulatory interactions arise where they are predicted to be similar, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

@@

\section{Methods}
\subsection{Gene regulatory network}
We represent gene regulatory networks as a matrix of weights, where rows index transcription factors (TFs) and columns index genes. The weight in a given position represents the regulatory weight of that TF on that gene; positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction. We begin with an existing framework for estimating the interaction weights in a gene regulatory network (GRN). In the Inferelator algorithm, linear differential equations are used to model changes in gene expression \cite{bonneau_inferelator:_2006-1} The rate at which the observed mRNA expression of gene $i$, $x_i$, changes is governed by degradation of existing transcripts with rate $\alpha$ plus a linear combination of transcription factor (TF) expressions. 
\begin{equation}
\frac{\mathrm d}{\mathrm d t} x_i = -\alpha_{i}x_{i} + \sum \beta_{i,j}x_{j}
\end{equation}
We are interested in learning $\beta$, the matrix defining the influence of each TF on each gene. We fix the decay rate $\alpha$ for all genes, and set it assuming a time-constant of 10 minutes \cite{hambraeus_genome-wide_2003, selinger_global_2003}, as in \cite{greenfield_robust_2013}. Let $x_i(t)$ be the expression of gene $i$ at time $t$. Given time-series data on the expression of gene $i$ at timepoints $t_k$ and $t_{k+1}$, we can approximate the rate of change of $x_i$ as $x_i'(t_k)=\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k}$. We can treat steady-state data as having a derivative of zero. This gives us, for each gene $i$ and time $t_{k}$ an equation

\begin{equation}
\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k} + \alpha_{i}x_{i}(t_k)= \sum \beta_{i,j}x_{j}(t_k)
\end{equation}
\noindent We can summarize these equations in matrix form as
\begin{equation}
Y = X \beta 
\end{equation}
\noindent We approach the learning of the $\beta$ matrix using regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the Inferelator algorithm, which we extend to the case of simultaneously inferring the GRNs of multiple species.

\subsection{Fused gene regulatory networks}
We hypothesize that related species should be governed by similar but not necessarily identical gene regulatory networks. We approach the problem of cross-species GRN inference by introducing constraints into the above regression formulation to penalize differences between interaction weights that are expected to be similar based on gene orthology. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $p_{\lambda_S}(\beta^A_{g,k} - \beta^B_{h,l})$ to the quantity being minimized in order to ensure similarity. The parameter $\lambda_S$ controls how highly to penalize differences between fused coefficients and controls the tradeoff between fitting the expression data and producing a set of networks that conform to evolutionary constraints. This gives us the final equation to be minimized 
\begin{equation}
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} p_{\lambda_S}(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})
\end{equation}

@@

When the penalty $p_{\lambda_S}$ is the L2 norm, the entire problem can be solved through linear regression with a suitable design matrix. The regulators of a pair of genes can be solved independently as long as the pair is not linked by fusion constraints, or a chain of fusion constraints; thus the problem of cross-species GRN inference decomposes into a large number of manageable regression problems as long as the orthology mappings are sparse.

When there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism), the choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). Part of the appeal of the approach, however, is that it allows pooling of data even when there is incomplete orthology. By introducing constraints on the similarity of individual interactions, rather than on the networks as a whole \cite{liu2011temporal}, we can pool some information across species even when an arbitrarily small fraction of genes have orthologs. This is particularly useful when dealing with a large number of species; pairwise orthologies may be nearly complete even when the number of genes present in every organism is small.

\subsection{Solving fused L2 problems using augmented matrices}
The problem of solving a multi-valued regression problem can be converted into the problem of solving a single valued regression problem by vectorization.
Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$. We solve

@@

%We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication

Given $X_1... X_S$ matrices, $X_i \in R^{N_i \times K_i}$ we define $N = \sum_{i=0}^S N_i$, $K = \sum_{i=0}^SK_i$ and $X \in R^{N \times K}$ by diagonally concatenating $X_1...X_S$.
In this new matrix, the first $K_1$ features correspond to features of $X_1$, the next $K_2$ features the features of $X_2$ etc.
Similarly, the first $N_1$ samples correspond to samples of $X_1$, the next $N_2$ to $X_2$ etc.
Solving the regression problem associated with each source is equivalent to solving the problem $\underset{W}{argmin}||XW-Y||$ where $Y$ is defined in the same way as $X$.
The weight matrix $W$ has a block diagonal structure:

@@

The fused multi-source regression problem can solved by constructing a multi-source design matrix $X$ and data matrix $Y$, and then vectorizing.
The regression problems associated with two columns $g,h$ of $Y$, where $g$ came from $Y_1$ and $h$ came from $Y_2$, are independent, unless there is a constraint chain of length $L$ with the form (where $x$ is arbitrary)

%$$(G, j_0, g, J_1, j_1, m_1) \rightarrow (J_1, j_1, m_1, J_2, j_2, m_2) \rightarrow (J_2,j_2,m_2,J_3,j_3,_m3)...(J_P, j_P, m_P, H, j_{P+1}, h)$$
$$(Y_1, j_1, g, Y_2, j_2, h) \rightarrow (Y_2, j_2, h, Y_3, j_3, k)$$

where the $j$s refer to transcription factors. In subproblem 1, $j_1$ regulates gene $g$; $j_1$ is orthologous to $j_2$, a transcription factor of subproblem 2, which regulates $h$, a gene which is orthologous to gene $g$. There exists a chain because ($j_1$, $g$) is a TF, gene pair which is fused with ($j_2$, $h$), which is fused to ($j_3$, $k$).
We use depth-first search to identify linked columns of $Y$, and then solve the fused regression problem directly by vectorizing those columns.
In order to incorporate the fusion penalty, we append to the resulting design matrix one row \emph{per connected constraint}, which contains $\lambda_S$ in the column associated with $Y_g,i,j,$ and $-\lambda_S$ to the column associated with $Y_h, i',j'$ for constraint $(Y_g,i,j,Y_h,i',j')$. Call this resulting matrix $X_P$.
The vector $Y$ is augmented by appending an appropriate number of zeros to produce $Y_P$. As a result, each entry in the output of $X_P b$ in the columns associated with each constraint contain $\lambda_S(b_{i,j} - b_{i',j'})$.
These entries each contribute  $\lambda_S||b_{i,j} - b_{i',j'}||^{2}$ to the squared error $||X_Pb - Y_P||_2^{~2}$

In the context of regression for GRN inference with multiple cell sources, each gene $j$ from source $1$ has constraints $\forall_{i \in \{TFs\}} (1, i, j, 2, i’, j’)$ where $i',j'$ are the same gene in a different cell line.
As a result, each gene has only one constraint and the chain has length $L=2$.
%In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $L\sum_{i=0}^S K_i \times L \sum_{i=0}^S K_i$ matrix, taking $O((L\sum_{i=0}^S K_i)^3)$.
In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $p\sum_{i=0}^S K_i \times p \sum_{i=0}^S K_i$ matrix where $p$ is the size of a column subset such that there is a constraint chain $L$ linking an element of each column to an element in every other column in that subset, taking $O((p\sum_{i=0}^S K_i)^3)$.
Although the sparse structure of $X_P$ makes this somewhat more tractable, it is still impractical for large $L$.

@@

When the number of genes linked by fusion constraints is relatively small, the augmented design matrix approach we have described has the advantage of being efficiently solvable with a closed form solution. However, in the general case the augmented system of equations may be too large to solve efficiently with standard methods. To address this limitation, we developed an iterative solver that uses coordinate-wise descent to solve for solutions corresponding to a sequence of values of fusion penalty weights; since our fused L2 method uses a convex and differentiable penalty function, this approach converges to a global minimizer. Although less efficient than the augmented design matrix approach we developed, the iterative solver has the advantage of computing a solution path for $\lambda_S$.

@@

\begin{equation}
\text{argmin}_{\beta^S(t) : S_i, S_j \in \text{species}} \displaystyle\sum_{S_i, S_j \in \text{species}} ||X^{S_i}\beta^{S_i}(t) - Y^{S_i}||^2 + \lambda_R||\beta^{S_i}(t)||^2 + \displaystyle \sum_{\beta_{g,k}^{S_i} \approx \beta_{h,l}^{S_j}} \lambda_S(\beta^{S_i}_{g,k}(t) - \beta_{h,l}^{S_j}(t-1)^2)
\end{equation}
Note that this is almost identical to equation 5, but now the network $\beta$ is a function of the iteration number $t$. On each step, the we compute $\beta$s that minimize a penalized cost function where the fusion penalty encourages similarity of a pair of parameters, using the previous iteration's solution. This process is iterated until the estimated $\betas$s converge. Because each iteration reduces the error between $\beta(t)$ and $\beta(t-1)$, and because $\beta(t) = \beta(t-1)$ is the globally optimal solution, this process must eventually converge to the same network as (cite 5). Although we have not produced bounds on the convergence rate, in practice a small number of iterations are necessary.

>>>>

\title{Fused regression for multi-source network inference}
\author{Kari Y. Lam, Zachary M. Westrick, Christian L. M\"{u}ller, Lionel Christiaen, Richard Bonneau}
\maketitle

\begin{abstract}
Understanding gene regulatory networks is critical to learning how cells regulate their behavior and respond to external stimuli. To this end, methods for global network inference have been developed and applied to a variety of species. However, most approaches consider the problem of network inference independently in each species, despite evidence that gene regulation is conserved even in distantly related species. We introduce a method for multi-source network inference that allows simultaneous estimation of gene regulatory networks in multiple species or biological processes through the incorporation of priors based on known gene relationships such as orthology. This approach improves network inference performance even when the orthology mapping in incomplete. We further refine this method with an algorithm that attempts to learn the true conserved subnetwork from a larger set of potentially conserved interactions. We demonstrate the utility of our method not only in cross species network inference, but also in merging data collected on different experimental platform and in incorporating binding similarity predictions.
\end{abstract}

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable \cite{bonneau_predictive_2007, ciofani_validated_2012, carro_transcriptional_2010}. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist adequate data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from existing studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning multiple networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

Moreover, as it becomes increasingly possible to learn genome-wide regulatory networks, we can begin to compare and to test whether there is conservation of networks across species and biological processes. Our use of model organisms to study biological processes and diseases relevant to humans relies on the assumption of conservation; yet this has not been tested at the genome scale.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference \cite{bonneau_predictive_2007, arrieta-ortiz_experimentally_2015}. The core of both methods is the observation that biological information about the relatedness of two networks can be used to constrain which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our first method -- fused ridge -- uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously learn the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different). With this approach, we seek to test the hypothesis of conservation of interactions across networks.

In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance \cite{satou2006gene, hinman2009evolution,tanay2005conservation,erwin2009evolution}. In our fused L2 approach, we make the assumption that if a pair of TFs is orthologous to one another, then their regulatory effects on a pair of orthologous genes should be similar. These orthology relationships form the basis of a set of constraints which favor -- but do not require -- networks in which orthologous transctiption factors regulate orthologous genes. As a result, data in one species can improve network inference performance in another species  (and vice versa). This general framework for multi-species network inference can be extended to an arbitrary number of distinct organisms, each contributing data with only partially overlapping sets of genes. This is an advantage over existing approaches to multi-species network inference, which infer only a sub-network for which orthologs exist in every species \cite{joshi_multi-species_2015}. This approach of introducting constraints on the similarity between specific regulatory interactions can be extended beyond the case of multi-species network-inference from orthology; any biological prior on similarity can be used in place of orthology. For example, we can introduce constraints that favor genes in the same operon or having similar binding sites, having similar regulators.

Existing multi species approaches often use orthology as a proxy for functional conservation \cite{penfold_inferring_2015, joshi_multi-species_2015, kashima_simultaneous_2009, zhang2010nearly}, or attempt to learn functional similarity via expression data \cite{gholami_cross-species_2010}. Orthology -- the measure of gene similarity we use -- can be approximated using readily identifiable sequence similarity. Sequence similarity is often a useful predictor of functional similarity \cite{wilson_assessing_2000}. In multi-species network inference, our fused L2 approach minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with evolutionary constraints created using orthology. However, we recognize that some genes may have taken on different functions and therefore may have new regulatory interactions. Gene duplications and neofunctionalization in multi-species network inference \cite{eisen_phylogenomics:_1998}, and changes in chromatin configuration in multi-cell line network inference \cite{li_role_2007}, are cases where enforcing fusion of constraints does not produce networks which reflect true differences in regulation between networks. These cases can be difficult to learn by simply comparing independently inferred networks; because network inference is typically underconstrainted \cite{marbach_revealing_2010-1}, there could be other plausible networks in which the coefficients being compared are similar. Our adaptive fusion method allows us to learn differences in networks predicted to be similar based on orthology or gene matching: because the constraints we impose favor weight configurations in which fused interaction weights are similar to one another, a solution in which a priori similar regulatory interactions are found to be different is strong evidence that they really are different. When differences in regulatory interactions arise where they are predicted to be similar, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

@@

\section{Methods}
\subsection{Gene regulatory network}
We model each gene's expression as a linear combination of transcription factors (TFs), and we seek to identify the TFs which regulate each gene, and the strength of the TF-gene interactions. In the Inferelator algorithm, linear differential equations are used to model changes in gene expression \cite{bonneau_inferelator:_2006-1} Our primary data for learning gene regulatory networks is expression data, time-series and steady state experiments. We use time-series data to approximate the rate of change of expression, and we treat steady-state experiments as conditions that have reached equilibrium. The rate at which $x_{i}$, the observed mRNA expression of gene $i$, changes, is governed by degradation of existing transcripts with rate $\alpha$ plus a linear combination of transcription factor (TF) expressions. 
\begin{equation}
\frac{\mathrm d}{\mathrm d t} x_i = -\alpha_{i}x_{i} + \sum \beta_{i,j}x_{j}
\end{equation}
where $\beta_{i,j}$ represents the weight of TF $j$ on gene $i$, and $\alpha$ is the decay rate of gene $i$. We fix the decay rate $\alpha$ for all genes, and set it assuming a time-constant of 10 minutes \cite{hambraeus_genome-wide_2003, selinger_global_2003}, as in \cite{greenfield_robust_2013}. Let $x_i(t)$ be the expression of gene $i$ at time $t$. Given time-series data on the expression of gene $i$ at timepoints $t_k$ and $t_{k+1}$, we can approximate the rate of change of $x_i$ as $x_i'(t_k)=\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k}$. We treat steady-state data as having a derivative of zero. This gives us, for each gene $i$ and time $t_{k}$ an equation

\begin{equation}
\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k} + \alpha_{i}x_{i}(t_k)= \sum \beta_{i,j}x_{j}(t_k)
\end{equation}
\noindent We can summarize these equations in matrix form as
\begin{equation}
Y = X \beta 
\end{equation}

We are interested in learning $\beta$, the matrix representation of the gene regulatory network, where the weight in a given position represents the regulatory weight of a TF on a gene. Positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction.

\noindent We approach the learning of the $\beta$ matrix using regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the Inferelator algorithm, which we extend to the case of simultaneously inferring the GRNs of multiple species.

\subsection{Fused gene regulatory networks}
We hypothesize that related species should be governed by similar but not necessarily identical gene regulatory networks. We approach the problem of cross-species GRN inference by introducing constraints into the above regression formulation to penalize differences between interaction weights that are expected to be similar based on, say, gene orthology. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $p_{\lambda_S}(\beta^A_{g,k} - \beta^B_{h,l})$ to the quantity being minimized in order to ensure similarity. The parameter $\lambda_S$ controls how highly to penalize differences between fused coefficients and controls the tradeoff between fitting the expression data and producing a set of networks that conform to evolutionary constraints. This gives us the final equation to be minimized 
\begin{equation}
%\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} p_{\lambda_S}(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\substack{(g,h) \in orth,\\
 (h,l) \in orth}}x p_{\lambda_S}(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})
\end{equation}

@@

\subsection{Fused L2 penalty for multi-source network inference}
When the penalty $p_{\lambda_S}$ is the L2 norm, as in our fused L2 approach:
\begin{equation}
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \lambda_S \sum_{\substack{(g,h) \in orth,\\
 (h,l) \in orth}} \vert \vert (\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l}) \vert \vert ^2
\end{equation}

the entire problem can be solved through linear regression with a suitable design matrix. The regulators of a pair of genes can be solved independently as long as the pair is not linked by fusion constraints, or a chain of fusion constraints; thus the problem of cross-species GRN inference decomposes into a large number of manageable regression problems as long as the orthology mappings are sparse.

In order to better understand this, consider the case where there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism). The choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). Part of the appeal of the approach, however, is that it allows pooling of data even when there is incomplete orthology. By introducing constraints on the similarity of individual interactions, rather than on the networks as a whole \cite{liu2011temporal}, we can pool some information across species even when an arbitrarily small fraction of genes have orthologs. This is particularly useful when dealing with a large number of species; pairwise orthologies may be nearly complete even when the number of genes present in every organism is small.

\subsection{Solving fused L2 problems using augmented matrices}
We use the fact that ridge regression can be solved by ordinary least squares regression on an augmented dataset, by appending a scaled identity matrix to the design matrix, and zeros to the response matrix. The problem of solving a multi-valued regression problem can also be converted into the problem of solving a single valued regression problem by vectorization.

Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$. We solve

@@

%We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication. In an ordinary ridge regression problem, this is unnecessary, as the columns of the weight matrix can be solved independently. However, in fused regression, information is shared between columns, and therefore they must be solved for simultaneously. If two entries are in the same fusion constraint, or if there exists a constraint chain containing entries from two columns, those columns must be solved together. Two entries may be fused if say, two TFs are orthologous, and two genes are orthologous--then, the interaction between the first TF, gene pair should be similar to the second TF, gene pair. A constraint chain can exist if there are complex orthology mappings, say, if there are multiple similar sequence signals, and three or more entries are fused.

%Given $X_1... X_S$ matrices, $X_i \in R^{N_i \times K_i}$ we define $N = \sum_{i=0}^S N_i$, $K = \sum_{i=0}^SK_i$ and $X \in R^{N \times K}$ by diagonally concatenating $X_1...X_S$.
%In this new matrix, the first $K_1$ features correspond to features of $X_1$, the next $K_2$ features the features of $X_2$ etc.
%Similarly, the first $N_1$ samples correspond to samples of $X_1$, the next $N_2$ to $X_2$ etc.
%Solving the regression problem associated with each source is equivalent to solving the problem $\underset{W}{argmin}||XW-Y||$ where $Y$ is defined in the same way as $X$.
%The weight matrix $W$ has a block diagonal structure:

@@

The fused multi-source regression problem can solved by constructing a multi-source design matrix $X$ and data matrix $Y$, and then vectorizing.
Suppose we have a constraint with the form

%The regression problems associated with two columns $A,A'$ of $Y$, where $A$ came from $Y_1$ and $A'$ came from $Y_2$, are independent, unless there is a constraint chain of length $L$ with the form

%$$(G, j_0, g, J_1, j_1, m_1) \rightarrow (J_1, j_1, m_1, J_2, j_2, m_2) \rightarrow (J_2,j_2,m_2,J_3,j_3,_m3)...(J_P, j_P, m_P, H, j_{P+1}, h)$$
%$$(Y_1, j_1, g, Y_2, j_2, h) \rightarrow (Y_2, j_2, h, Y_3, j_3, k)$$
$((\beta_1, A, B), (\beta_1, A, C), (\beta_2, A', B'))$

Here, we have two datasets, which we refer to as 1 and 2. We have information about the similarity of TF A in network 1 to TF A' in network 2, and about the similarity of genes B and C in network 1 to gene B' in network 2. We expect, therefore, that if an interaction exists between A and B or A and C in the first network, that these interactions be similar to the interaction between A' and B' in the second network, and we penalize solutions which do not have this feature.

To find all such linked problems, we use depth-first search to identify linked columns of the TF expression matrix $Y$, then solve the fused regression problem directly by vectorizing corresponding columns. To incorporate the fusion penalty, we append to the resulting block diagonal design matrix one row per connected constraint, which contains $\lambda_S$ in the column associated with the TF, gene pair (A, B) from dataset 1, and $-\lambda_S$ to the column associated with TF, gene pair (A', B') from dataset 2, as well as another one for (A, C), (A', B'). The vector $Y$ is augmented by appending an appropriate number of zeros. This results in the addition of the penalties, $\lambda_S||b_{A,B} - b_{A',B'}||^{2}$ and $\lambda_S||b_{A,C} - b_{A',B'}||^{2}$ to the squared error $||AW-y||^2$.

In most cases, we have found the direct solve using augmented matrices to be adequate. However, there are certainly use cases of fused regression where solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $p\sum_{i=0}^S K_i \times p \sum_{i=0}^S K_i$ matrix where $K_i$ is the number of features in each network and $p$ is the size of a column subset such that there is a constraint chain linking an element of each column to an element in every other column in that subset, taking $O((p\sum_{i=0}^S K_i)^3)$.
Although the sparse structure of $X_P$ makes this somewhat more tractable, it is still impractical for large chains.

%where the $j$s refer to transcription factors. In subproblem 1, $j_1$ regulates gene $g$; $j_1$ is orthologous to $j_2$, a transcription factor of subproblem 2, which regulates $h$, a gene which is orthologous to gene $g$. There exists a chain because ($j_1$, $g$) is a TF, gene pair which is fused with ($j_2$, $h$), which is fused to ($j_3$, $k$).
%We use depth-first search to identify linked columns of $Y$, and then solve the fused regression problem directly by vectorizing those columns.
%In order to incorporate the fusion penalty, we append to the resulting design matrix one row \emph{per connected constraint}, which contains $\lambda_S$ in the column associated with $Y_g,i,j,$ and $-\lambda_S$ to the column associated with $Y_h, i',j'$ for constraint $(Y_g,i,j,Y_h,i',j')$. Call this resulting matrix $X_P$.
%The vector $Y$ is augmented by appending an appropriate number of zeros to produce $Y_P$. As a result, each entry in the output of $X_P b$ in the columns associated with each constraint contain $\lambda_S(b_{i,j} - b_{i',j'})$.
%These entries each contribute  $\lambda_S||b_{i,j} - b_{i',j'}||^{2}$ to the squared error $||X_Pb - Y_P||_2^{~2}$

%In the context of regression for GRN inference with multiple cell sources, each gene $j$ from source $1$ has constraints $\forall_{i \in \{TFs\}} (1, i, j, 2, i’, j’)$ where $i',j'$ are the same gene in a different cell line.
%As a result, each gene has only one constraint and the chain has length $L=2$.
%In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $L\sum_{i=0}^S K_i \times L \sum_{i=0}^S K_i$ matrix, taking $O((L\sum_{i=0}^S K_i)^3)$.
%In general, however, solving the fused regression problem directly, even after factoring constraints, involves at least one inversion of an $p\sum_{i=0}^S K_i \times p \sum_{i=0}^S K_i$ matrix where $p$ is the size of a column subset such that there is a constraint chain $L$ linking an element of each column to an element in every other column in that subset, taking $O((p\sum_{i=0}^S K_i)^3)$.
%Although the sparse structure of $X_P$ makes this somewhat more tractable, it is still impractical for large $L$.

@@

When the number of genes in each fusion constraints is relatively small, the augmented design matrix approach we have described has the advantage of being efficiently solvable with a closed form solution. However, in some cases the augmented system of equations may be too large to solve efficiently with standard methods. To address this limitation, we developed an iterative solver that uses coordinate-wise descent to solve for solutions corresponding to a sequence of values of fusion penalty weights; since our fused L2 method uses a convex and differentiable penalty function, this approach converges to a global minimizer. Although less efficient than the augmented design matrix approach we developed, the iterative solver has the advantage of computing a solution path for $\lambda_S$.

@@

\begin{equation}
\text{argmin}_{\beta^{S_i(t)}, \beta^{S_j(t)}} \displaystyle\sum_{S_i, S_j \in \text{species}} ||X^{S_i}\beta^{S_i}(t) - Y^{S_i}||^2 + \lambda_R||\beta^{S_i}(t)||^2 + \displaystyle \sum_{\beta_{g,k}^{S_i} \approx \beta_{h,l}^{S_j}} \lambda_S(\beta^{S_i}_{g,k}(t) - \beta_{h,l}^{S_j}(t-1)^2)
\end{equation}
Note that this is almost identical to equation 5, but now the network $\beta$ is a function of the iteration number $t$. On each step, the we compute $\beta$s that minimize a penalized cost function where the fusion penalty encourages similarity of a pair of parameters, using the previous iteration's solution. This process is iterated until the estimated $\betas$s converge. Because each iteration reduces the error between $\beta(t)$ and $\beta(t-1)$, and because $\beta(t) = \beta(t-1)$ is the globally optimal solution, this process must eventually converge to the same network as (cite 5). Although we have not produced bounds on the convergence rate, in practice a small number of iterations are necessary.
